---
format: 
  html:
    toc: true 
editor: visual
execute:
  warning: false
  message: false
---

## Objectives: quantifying fire/disease impacts   
- Quantify biomass, stem density changes pre to post fire. Compare across burn scar, species, and status.   
  - Used mixed effects models to understand underlying relationships.   
- Track stems through time to quantify rates of mortality, snag fall and CWD decomposition. 
  - Use survival analysis (mortality), ?? (snag fall), regression (CWD decomp.)    

### Needed libraries   
```{r}
library(tidyverse)
library(broom)
library(gtsummary)
library(survival)
library(ggsurvfit)
library(lme4) 
library(lmerTest)
library(here)
library(knitr)
library(patchwork)
library(gt)
library(kableExtra)
```    

### Recycled functions  
```{r}
read_me <- function(path){
  read_csv(here("data", path))
}
export_image <- function(img, path){
  ggsave(filename = here("exports", path), plot=img, width=12, height=8)
}

climate_theme <- function(color = "slategray1") {
  theme(
  plot.title = element_text(
      size = 16,
      family = "Times New Roman",
      color = "black", 
      margin = margin(b = 15)
    ),
    # add border 1)
    panel.border = element_rect(colour = "black", fill = NA, linetype = 1),
    # color background 2)
    panel.background = element_rect(fill = color),
    # modify grid 3)
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y =  element_blank(),
    panel.grid.minor.y = element_blank(),
    # modify text, axis and colour 4) and 5)
    axis.text = element_text(colour = "black",  family = "Times New Roman"),
    axis.title = element_text(colour = "black", family = "Times New Roman"),
    axis.ticks = element_line(colour = "black"),
    legend.position = "right"
  )
}
# darkolivegreen3

```

## Univariate statistics of fire and disease impacts 

### Living Biomass changes   

```{r}
#| fig-width: 14
#| fig-height: 16


biomass_changes <- read_me("./clean/post_fire_summary.csv")
burn_change <- function(burn, time1) {
  df <- biomass_changes |>
    filter(BurnScar == burn) |>
    select(!c(Elevation,Slope,Fire)) |>
    mutate(Pre = case_when(
      SampleYear <= time1 ~ "Pre",
      SampleYear >=  time1 ~ "Post",
      TRUE ~ "NA")) 
  df_mod1 <- df |>
    filter(Pre %in% c("Pre","Post")) |>
    group_by(Plot, Species, BurnScar, PRam, Pre) |>
    summarise(
      TotalStem   = sum(TotalStem, na.rm = TRUE),
      LiveBiomass = sum(LiveBiomass, na.rm = TRUE),
      .groups = "drop"
    ) |>
    pivot_wider(names_from="Pre", values_from=c("TotalStem","LiveBiomass"))
  # only want to evaluate data with both pre and post 
  df_mod2 <- na.omit(df_mod1)
  df_mod3 <- df_mod2 |>
    mutate(ChangeStem = TotalStem_Post - TotalStem_Pre,
      ChangeBiomass = LiveBiomass_Post - LiveBiomass_Pre) |>
    select(c(1,2,3,4,9,10))
  return(df_mod3)
}
# apply summary function to dataset we want
sites_tree <- tibble(
  burn  = c("dolan", "sober", "chalk", "control", "dolan_chalk", "sober_chalk"),
  time1 = c(2020, 2016, 2007, 2016, 2007, 2007)
)
demo_tree <- pmap_df(sites_tree, burn_change)

post_fire_summary_biomass <- demo_tree |>
  group_by(Species, BurnScar, PRam) |>
  summarize( n = n(),
            MeanStemChange = mean(ChangeStem, na.rm=TRUE),
            MeanBiomassChange = mean(ChangeBiomass, na.rm=TRUE),
            MedianStemChange    = median(ChangeStem, na.rm = TRUE),
            MedianBiomassChange = median(ChangeBiomass, na.rm = TRUE),
            SdStemChange = sd(ChangeStem, na.rm=TRUE),
            SdBiomassChange = sd(ChangeBiomass, na.rm=TRUE)
) |>
  rename(
         `Mean Change in Stem Density` = 3,
     `Mean Change in Biomass` = 4,
     `Median Change in Stem Density` = 5,
     `Median Change in Biomass` = 6,
  `SD Change in Stem Density` = 7,
  `SD Change in Biomass` = 8)


kable(post_fire_summary_biomass)


den <- demo_tree |>
    ggplot(aes(x=ChangeStem, y= ChangeBiomass, color = Species, linetype = as.factor(PRam))) +
    geom_point() +
    geom_smooth(method = "lm", se =FALSE) +
    climate_theme() 

dis <- demo_tree |>
    ggplot(aes(x=BurnScar, y= ChangeBiomass, color = BurnScar, fill = as.factor(PRam))) +
    geom_boxplot() +
    geom_jitter(shape = 21, 
                position = position_jitterdodge(0.8)) + 
    climate_theme() +
    scale_color_viridis_d()

dis2 <- demo_tree |>
    ggplot(aes(x=BurnScar, y= ChangeStem, color = BurnScar, fill = as.factor(PRam))) +
    geom_boxplot() +
    geom_jitter(shape = 21, 
                position = position_jitterdodge(0.8)) + 
    climate_theme()
dis3 <- demo_tree  |>
    ggplot(aes(x=Species, y= ChangeBiomass, color = Species, fill = as.factor(PRam))) +
    geom_boxplot() +
    geom_jitter(shape = 21, 
                position = position_jitterdodge(0.8)) + 
    climate_theme()

dis4 <- demo_tree |>
    ggplot(aes(x=Species, y= ChangeStem, color = Species, fill = as.factor(PRam))) +
    geom_boxplot() +
    geom_jitter(shape = 21, 
                position = position_jitterdodge(0.8)) + 
    climate_theme()


(dis + dis2) / (dis3 + dis4) / den +
  plot_annotation(title = "Pre to Post Fire Living Tree Summary") +
plot_layout(heights = c(1, 1, 2)) 


```

## Underlying drivers of biomass and stem density changes   

### Mixed effects models   
-- Multiple regression with mixed effects models iterated across species     
    - Attempt to understand drivers of biomass changes. Is disease and fire interacting to reduce biomass? Is it even across species.       
- Improvements: Run multiple models per species with different explanatory variables, take weighted AIC and weighted effects.  
**Questions I hope to answer**  
Which features, topographic or epidemiological, are driving biomass changes in Redwood forests?  
Is there a linear relationship between the fixed explanatory variables and biomass?  

-- P-values for fixed variables below the 0.05 threshold will allow us to reject our null hypothesis and indicate a linear relationship between a fixed explanatory variable and biomass.



#### Plot biomass changes 

- Scaling values standardizes units? Worth doing? 

**Interactions**    
```{r}
#| fig-width: 12 
#| fig-height: 14

biomass <-  read_me("clean/livebiomass_mixedmodel.csv")

int_me <- function(df,x,y,int) {
  df |>
    ggplot(aes(x={{x}},y={{y}},color=as.factor({{int}}),group=as.factor({{int}}))) +
      geom_point() +
      geom_smooth(method="lm") +
      climate_theme()
}

p1 <- int_me(biomass,DeadHostBiomass, HostCwdMass, PRam)
p2 <- int_me(biomass,DeadHostBiomass, Elevation, PRam)
p3 <- int_me(biomass,Fire, LiveBiomass,PRam)
p4 <- int_me(biomass,DeadHostBiomass, Slope, Fire)


(p1 + p2) / (p3 + p4)


```

**Other interactions to graph?**    

**Adding time since fire,UMCA stem density**    
```{r}

biomass1 <- biomass |>
  mutate(tFire = case_when(
                          BurnScar == "chalk" ~ SampleYear - 2008,
                          BurnScar == "sober" ~ SampleYear - 2016,
                          BurnScar == "dolan" ~ SampleYear - 2020,
                          BurnScar == "dolan_chalk" ~ SampleYear - 2020,
                          BurnScar == "sober_chalk" ~ SampleYear - 2016,
                          BurnScar == "control" ~ 0 
)) 


umca <- biomass |>
  filter(Species == "UMCA") |>
  select(Plot, SampleYear, TotalLivingStem) |>
  rename(UmcaStemDensity = TotalLivingStem)

biomass2 <- biomass1 |>
  left_join(umca, by = c("Plot", "SampleYear"), relationship = "many-to-many")

```

#### Mixed Models     

The (1|SampleYear) notation is being rendered as (1&#124;SampleYear).   

```{r}
species <- biomass2 |>
  filter(Species %in% c('SESE', 'LIDE', 'QUAG', 'UMCA')) 

species_df <- unique(species$Species)

formulas <- c(
              "LiveBiomass ~ PlotCwdMass + HostCwdMass + DeadHostBiomass + Elevation + Slope + Fire + PRam + tFire + 
              UmcaStemDensity + (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass + DeadHostBiomass + Elevation + Slope + Fire + PRam + tFire + 
              (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass + DeadHostBiomass + Elevation + Slope + Fire + PRam + tFire + 
              UmcaStemDensity + SampleYear + (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass + DeadHostBiomass + Elevation + Slope + Fire + PRam + tFire + 
              UmcaStemDensity + (1|tFire) + (1|Plot)",
            "LiveBiomass ~ Fire * PRam * DeadHostBiomass * HostCwdMass * tFire + PlotCwdMass + Elevation + Slope +
                    (1|SampleYear) + (1|Plot)",
            "LiveBiomass ~ PlotCwdMass * HostCwdMass * DeadHostBiomass * Elevation * Slope * Fire * PRam +
                    (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass * HostCwdMass * DeadHostBiomass * Elevation * Slope * Fire * PRam * tFire +
                    (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass * HostCwdMass * DeadHostBiomass * Elevation * Slope * Fire * PRam * tFire +
                    (1|tFire) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass * DeadHostBiomass * Fire * PRam + Elevation * Slope + tFire +
                    (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass * DeadHostBiomass * Fire * PRam + Elevation * Slope +
                    (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass + DeadHostBiomass + Elevation + Slope + Fire + PRam + tFire +
                    (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass + DeadHostBiomass + Elevation + Slope + Fire + PRam + tFire +
                    (1|tFire) + (1|Plot)",
               "LiveBiomass ~ PlotCwdMass * HostCwdMass * DeadHostBiomass * Elevation * Slope * Fire * PRam +
                    (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass * DeadHostBiomass * Fire * PRam + Elevation * Slope +
                    (1|SampleYear) + (1|Plot)",
              "LiveBiomass ~ PlotCwdMass + HostCwdMass * DeadHostBiomass * Fire * PRam + Elevation * Slope +
                    (1|SampleYear) + (1|Plot)"
)
big_biomass_model <- function(species, form){
  df1 <- biomass2 |>
    filter(Species == {{species}}) |>
    mutate(across(c(PlotCwdMass, HostCwdMass, DeadHostBiomass, 
                  Elevation, Slope,LiveBiomass), scale))
  
  Equation <- as.formula(form) 
  tryCatch({
    m1 <- lmer(Equation, data=df1) 
    summary <- summary(m1)
    df_summary <- as.data.frame(summary$coefficients) |>
      mutate( Equation = form,
             Species = species,
              AIC = AIC(m1),
              RMSE = sqrt(mean(residuals(m1) ^ 2))) 
    df_summary$Term = rownames(df_summary)
    return(df_summary)
  }, error = function(e) {
    message("Error", e$message)
    return(NULL)
  })
}

#cleaning function for streamlined results
clean_me <- function(df) {
  aic_lookup <- df |> distinct(Species, AIC, RMSE, Equation)
  df |>
    rename(p = 5) |>
    mutate(
           Results = ifelse(
      p < 0.05,
      paste0("**", round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")**"),
      paste0(round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")")
    )) |>
    select(Species, Equation, Term, Results) |>
    distinct(Species, Equation, Term, .keep_all = TRUE) |>
    pivot_wider(names_from = "Term", values_from = "Results") |>
    left_join(aic_lookup, by = c("Species", "Equation")) |>
    select(Equation, Species, AIC, RMSE, everything())
}

add_models <- map_df(species_df, function(sp) {
  map_df(formulas, function(fm) {
    big_biomass_model(sp, fm)
  })
}) |>
  clean_me() |>
  group_by(Species) |>
  slice_min(AIC, n=5) 

add_models |>
  kable() |>
  kable_styling(fixed_thead = TRUE,
                bootstrap_options = c("striped", "hover")) |>
  row_spec(0, background = "black", color = "white") |>
  scroll_box(height = "600px", width = "100%")


```








