---
format: 
  html:
    toc: true 
editor: visual
execute:
  warning: false
  message: false
---


## Install Data Package   
We have created an R package, bigsurrr, for easy connection to the Big Sur dataset. These datasets are based off the original dataset and include computations beyond the original data. These assumptions and methods are outlined in the original dataset cleaning section.       

```{r}
#| code-fold: false
#| eval: false
install.packages("devtools")
pak::pak("benparker9/bigsurrr")
```

```{r}
#| code-fold: false
library(bigsurrr)
read_me()
```


### Data Source   
- Originally from Access database   
  - [Click here](https://github.com/benparker9/bigsur/tree/master/data/access) to see original data   



### Original Dataset Cleaning   
```{r}
library(tidyverse)
library(here)
library(readxl)
library(arrow)
library(sf)

describe_me <- function(df){
  cat("==Dimensions==\n")
  print(dim(df))
  cat("==Sampling Years Available==\n")
  print(sort(unique(df$SampleYear)))
  cat("==Column Names==\n")
  print(names(df))
  cat("==Quantitative Summary Insights==\n")
  quant_summary <- df |>
    select(where(is.numeric))
  print(summary(quant_summary))
#  cat("==Qualitative Summary Insights==\n")
#  qual_summary <- df |>
#    select(where(is.factor)) |>
#    count(across(everything())) 
#  print(qual_summary)
}
```    

### Original Datasets  

#### Plot Database    
- two datasets provided by the access database (plot, plot2)    
- one provided by this [queried excel sheet](https://docs.google.com/spreadsheets/d/1v8svlbehpm1javfmewqujvsj88bfvci0/edit?gid=250385454#gid=250385454) that was then [analyzed](https://benparker9.github.io/modeling-documents/cal_poly/sod-analysis.html) for burn scar, *P.ram* and fire status.    

```{r}
# Plot DB Extract, Transform
plot <- read_parquet("./data/arrow/plot1.parquet")
plot2 <- read_parquet("./data/arrow/plot2.parquet")
plotDB <- read_parquet("./data/arrow/plotDB.parquet")


# To test random sampling
plot_mod1 <- plotDB |>
  group_by(Plot) |>
  summarize(TotalTimesInventoried = n()) |>
  right_join(plotDB, by=c("Plot")) |>
  mutate(VisitFrequency = case_when(
                                    TotalTimesInventoried <= 3 ~ "Low Visit",
                                    TotalTimesInventoried > 3 & TotalTimesInventoried < 6 ~ "Mid Visit",
                                    TotalTimesInventoried >= 6 & TotalTimesInventoried < 9 ~ "Mid-High Visit",
                                    TRUE ~ "High Visit"),
         Radius = case_when(
                            VisitFrequency =="Low Visit"~1,
                            VisitFrequency =="Mid Visit"~ 3,
                            VisitFrequency =="Mid-High Visit"~ 6,
                            VisitFrequency =="High Visit"~ 10
))

lat_lon <- plot_mod1 |>
    select(Plot,Easting,Northing)

points_sf <- st_as_sf(lat_lon,
                      coords = c("Easting", "Northing"),
                      crs = 32610)  # UTM 10N - Based on big sur region 

# Convert to lat/long WGS 1984
points_wgs84 <- unique(st_transform(points_sf, crs = 4326))
# Join to plot frequency data 
plot_mod_map <- plot_mod1 |>
  select(Plot,TotalTimesInventoried,VisitFrequency,Radius) |>
  distinct(Plot, .keep_all = TRUE) |> #ai help
  right_join(points_wgs84, by="Plot") 
# Clean lat/lon from sf output to two columns with distinct lat/lon
plot_mod_map1 <- plot_mod_map |>
 separate(geometry, into = c("Lon", "Lat"), sep = " ") |>
 mutate(Lon = str_remove_all(Lon, "c\\(|,"),
 Lat = str_remove_all(Lat,"\\)"))
# clean up data types
plot_mod_map2 <- plot_mod_map1 |>
  mutate(Lon = as.numeric(Lon),
         Lat = as.numeric(Lat))


# Clean plot attributes db with no coordinate data -- will be used with tree,cwd data
plot_mod2 <- plot_mod1 |>
  select(!(c(Easting, Northing,VisitFrequency,Radius,TotalTimesInventoried)))

describe_me(plotDB)
```

#### Tree Database    
- Three datasets provided by the access database (tree1, tree2, tree3)    
- One provided by this [queried excel sheet](https://docs.google.com/spreadsheets/d/1HnFcAt-QIB9nOsEm94CMsk8XmoRtvoFf/edit?gid=1239576248#gid=1239576248) 


```{r}
# 116k total, 114k distinct, from google sheets
treeDB <- read_parquet("./data/arrow/tree_basic.parquet")
# 113k total and distinct
tree1 <- read_parquet("./data/arrow/tree_stems.parquet") |>
  select(StemSampleID, TreeSampleID, StemID, StemNumber, DBH, Status, DeadClass, LogDecayClass, SnagHeight,
         StumpHeight, StumpDiameter, StumpDecay, Prostrate, Canker)
# error mismatch of cols, tree2 only has TreeID and TreeNumber, but only way to get species
tree2 <- read_parquet("./data/arrow/tree_trees.parquet")
# can get other health columns merge on treeSampleID and treeID
tree3 <- read_parquet("./data/arrow/tree_samples.parquet")

# Need to quantify biomass, use allometric equations to convert DBH into biomass (kg C)
#allometric equations https://www.fs.usda.gov/nrs/pubs/jrnl/2014/nrs_2014_chojnacky_001.pdf
# equation is biomass(kg) = exp(b0 + b1 * log(DBH)) 
# parameters can be found on page 140 
eq <- function(b0,b1,DBH){
  exp(b0+b1*log(DBH))
}


# Only plan on studying Redwood, tanoak, bay laurel, ponderosa pine, douglas fir, madrone
tree_final <- treeDB |> 
  mutate(Species = str_replace(Species,"quag", "QUAG")) |>
  filter(Species %in% c("SESE","LIDE","UMCA","PIPO","PSME", "ARME", "QUAG")) |>
  mutate(Biomass = case_when(
    Species == "LIDE" ~ eq(-2.2198, 2.4410, DBH),
    Species == "SESE" ~ eq(-2.7765,2.4195, DBH),
    Species == "QUAG" ~ eq(-2.2198, 2.4410, DBH),
    Species == "UMCA" ~ eq(-2.2118, 2.4133, DBH),
    Species == "ARME" ~ eq(-2.2118, 2.4133, DBH),
    Species == "PSME" ~ eq(-2.4623, 2.4852, DBH),
    Species == "PIPO" ~ eq(-3.2007, 2.5339, DBH)
),
    Mortality = ifelse(Status  == "L", 0, 1)) 

describe_me(tree_final)

```



#### CWD Database   

- Two datasets provided by the access database (cwd, cwd_samples)   
- One dataset provided by Dr.Cobb (cwd_density)      

```{r}
# CWD Database 
cwd <- read_parquet("./data/arrow/cwd.parquet") |>
  rename(CwdId = CWD_ID,
         Plot = BSPlotNumber,
         Species = PlantSpeciesAcronym) |>
  select(CwdId,Plot,Species)

cwd_samples <- read_parquet("./data/arrow/cwd_samples.parquet") |>
  rename(CwdId = CWD_ID) |>
  select(CwdId,SampleYear,DecayClass,Volume)

# merge on CwdId to add plot,species id's
# these id's are unique to stems
cwd_final <- cwd_samples |>
  left_join(cwd, by="CwdId")

# Densities per species per decay class - experimental dataset provided by my professor
cwd_density <- read_parquet("./data/arrow/cwd_density.parquet") |>
  pivot_longer(cols=X1:X5, values_to="Density", names_to="DecayClass") |>
  rename(Species = Speices) |>
  mutate(DecayClass = str_remove(DecayClass, "X"))

# need to get overall average density per decay class for cwd data not studied
avg_decay <- cwd_density |>
  group_by(DecayClass) |>
  summarize(Density = mean(Density))

# merge for one cohesive density table to be merged
cwd_density2 <- bind_rows(cwd_density,avg_decay)

# Join with raw cwd data - only evaluating DecayClass 1-5 due to density data gaps (only 1-5 studied)
cwd_final2 <- cwd_final |>
  left_join(cwd_density2, by=c("Species","DecayClass")) |>
  filter(DecayClass <= 5)

# no na's dataset, need to add average density per decay class to the other species
cwd_no_na <- na.omit(cwd_final2)
na <- cwd_final2 |>
  filter(is.na(Density)) |>
  filter(!is.na(Species)) |>
  select(!Density)
na_mod1 <- na |>
  left_join(avg_decay, by="DecayClass")
# Na_mod1 is then ready to merge with cwd final
cwd_final3 <- bind_rows(cwd_no_na, na_mod1) 

# Need to calculate mass and scale to per hectare (plots are 1/20ha so scale by 20 to get ha)
# Mass (Mg) = volume (m3) * density (g/cm3)  * 20 
# Only to follow repeatedly measured cwd stems - at least measured twice 
# Only need id, sampleyear, DecayClass,Plot,Species,Mass
# Need to create t column where t0 is the first time the CwdId was measured 
# need to create column named y which is log(Mt / M0) 
cwd_final4 <- cwd_final3 |>
  mutate(Mass = (Volume * (Density)),
  DecayClass = as.numeric(DecayClass)) |>
  group_by(CwdId) |>
  filter(n() > 1) |>
  group_by(CwdId, Plot, DecayClass) |>
  mutate(
    tInit = min(SampleYear),
    t = SampleYear - tInit,
    Mass0 = Mass[t == 0],
    Y = log(Mass / Mass0)
  ) |>
  ungroup() 

describe_me(cwd_final4)
```

