---
format: 
  html:
    toc: true 
editor: visual
execute:
  warning: false
  message: false
---


## Install Data Package   
We have created an R package, bigsurrr, for easy connection to the Big Sur dataset. These datasets are based off the original dataset and include computations beyond the original data. These assumptions and methods are outlined in the following dataset cleaning.       

```{r}
#| code-fold: false
#| eval: false
install.packages("devtools")
pak::pak("benparker9/bigsurrr")
```

```{r}
#| code-fold: false
library(devtools)
library(bigsurrr)
read_me()
```


### Data Source   
- Originally from Access database   
  - [Click here](https://github.com/benparker9/bigsur/tree/master/data/access) to see original data   



### Plot Database    
- two datasets provided by the access database (plot, plot2)    
- one provided by this [queried excel sheet](https://docs.google.com/spreadsheets/d/1v8svlbehpm1javfmewqujvsj88bfvci0/edit?gid=250385454#gid=250385454) that was then [analyzed](https://benparker9.github.io/modeling-documents/cal_poly/sod-analysis.html) for burn scar, *P.ram* and fire status.    

```{r}
library(tidyverse)
library(arrow)
library(sf)
# Plot DB Extract, Transform
plot <- read_parquet("./data/arrow/plot1.parquet")
plot2 <- read_parquet("./data/arrow/plot2.parquet")
plotDB <- read_parquet("./data/arrow/plotDB.parquet")


# To test random sampling
plot_mod1 <- plotDB |>
  group_by(Plot) |>
  summarize(TotalTimesInventoried = n()) |>
  right_join(plotDB, by=c("Plot")) |>
  mutate(VisitFrequency = case_when(
                                    TotalTimesInventoried <= 3 ~ "Low Visit",
                                    TotalTimesInventoried > 3 & TotalTimesInventoried < 6 ~ "Mid Visit",
                                    TotalTimesInventoried >= 6 & TotalTimesInventoried < 9 ~ "Mid-High Visit",
                                    TRUE ~ "High Visit"),
         Radius = case_when(
                            VisitFrequency =="Low Visit"~1,
                            VisitFrequency =="Mid Visit"~ 3,
                            VisitFrequency =="Mid-High Visit"~ 6,
                            VisitFrequency =="High Visit"~ 10
))

lat_lon <- plot_mod1 |>
    select(Plot,Easting,Northing)

points_sf <- st_as_sf(lat_lon,
                      coords = c("Easting", "Northing"),
                      crs = 32610)  # UTM 10N - Based on big sur region 

# Convert to lat/long WGS 1984
points_wgs84 <- unique(st_transform(points_sf, crs = 4326))
# Join to plot frequency data 
plot_mod_map <- plot_mod1 |>
  select(Plot,TotalTimesInventoried,VisitFrequency,Radius) |>
  distinct(Plot, .keep_all = TRUE) |> #ai help
  right_join(points_wgs84, by="Plot") 
# Clean lat/lon from sf output to two columns with distinct lat/lon
plot_mod_map1 <- plot_mod_map |>
 separate(geometry, into = c("Lon", "Lat"), sep = " ") |>
 mutate(Lon = str_remove_all(Lon, "c\\(|,"),
 Lat = str_remove_all(Lat,"\\)"))
# clean up data types
plot_mod_map2 <- plot_mod_map1 |>
  mutate(Lon = as.numeric(Lon),
         Lat = as.numeric(Lat))

# exported this for monte carlo simulation
```

#### State factors, tFire and final cleaning      

<a href=./state-factors.html target = "_blank">Extraction script</a>  

```{r}
#plot data
#CePlot 0=has ever had veg cov class <10%; 1= >10%; 2= >25%)
raw_plot <- read_parquet('./data/arrow/raw_plotDB.parquet') |>
  slice(-1) |>
  select(1,2, 5, 27, 40, 50, 51, 63, 64) |>
  rename(PlotBS = 2,
         ForestType = 3,
         YearsBurned = 4,
         CePlot = 5,
         Trail = 6,
         Owner = 7,
         PlotAspect = 8,
         OverallAspect = 9) |>
  mutate(YearsBurned = str_replace_all(YearsBurned, ",", " ")) |>
  separate_wider_delim(YearsBurned, delim = " ", names_sep = "_", 
                       too_few = "align_start") |>
  pivot_longer(cols = YearsBurned_1:YearsBurned_9, names_to = "Year", 
               values_to = "YearBurned") |>
  mutate(YearBurned = str_replace_all(YearBurned, "none", NA_character_)) |>
  drop_na(YearBurned) |>
  filter(!YearBurned == "") |>
  distinct()

# time since fire calculations
tFire <- raw_plot |> select(PlotBS,YearBurned) |> rename(Plot = 1) 

plot <- read_parquet('./data/arrow/plotDB.parquet')

p <- plot |> select(Plot, SampleYear, BurnScar) 

r <- left_join(p, tFire, by = "Plot") 

r2 <- r |>
  group_by(Plot, SampleYear, BurnScar) |>
  summarize(YearBurnedThirdMax = nth(sort(YearBurned, decreasing = TRUE), 3),
            YearBurnedSecondMax = nth(sort(YearBurned, decreasing = TRUE), 2),
            YearBurnedMax = max(YearBurned)) |>
  mutate(across(c(YearBurnedThirdMax, YearBurnedSecondMax, YearBurnedMax), as.numeric))

t <-r2 |>
  filter(BurnScar == "sober" & SampleYear > 2016  ) 

## time since fire calcs
# in control cat, 2013 fire? 
r3 <- r2 |>
  mutate(tFire = case_when(
                           BurnScar == "dolan" & SampleYear >= 2020 ~ SampleYear - 2020,
                           BurnScar == "dolan" & SampleYear < 2020 ~ SampleYear - YearBurnedMax,
                           BurnScar == "dolan_chalk" & SampleYear >= 2020 ~ SampleYear - 2020,
                           BurnScar == "dolan_chalk" & SampleYear < 2020 & SampleYear >= 2008 ~ SampleYear - 2008,
                          BurnScar == "dolan_chalk" & SampleYear < 2008 ~ SampleYear - YearBurnedSecondMax,
                           BurnScar == "sober" & SampleYear > 2016 ~ SampleYear - 2016,
                          BurnScar == "sober" & SampleYear < 2016 ~ SampleYear - YearBurnedSecondMax,
                           BurnScar == "sober_chalk" & SampleYear > 2016 ~ SampleYear - 2016,
                           BurnScar == "sober_chalk" & SampleYear < 2016 & SampleYear > 2008 ~ SampleYear - 2008,
                            BurnScar == "sober_chalk" & SampleYear < 2008 ~ SampleYear - YearBurnedThirdMax,
                            BurnScar == "chalk" & SampleYear > 2008 ~ SampleYear - 2008,
                          BurnScar == "chalk" & SampleYear < 2008 ~ SampleYear - YearBurnedSecondMax,
                          BurnScar == "control" & SampleYear >= 2013 & YearBurnedMax == 2013 ~ SampleYear - YearBurnedMax,
                          BurnScar == "control" & SampleYear < 2013 & YearBurnedMax == 2013 ~ SampleYear - YearBurnedSecondMax,
                          BurnScar == "control" & SampleYear >= 2006 & !is.na(YearBurnedMax)  ~ SampleYear - YearBurnedMax)) |>
    select(Plot, BurnScar, SampleYear, tFire, YearBurnedMax, YearBurnedSecondMax, YearBurnedThirdMax) 

#r3 can be audited for logic, 280 plots w/ sampleyear and time since fire
# 92 plots with no tFire Data - 
# these are cases where SampleYear < only 1 yearBurned datapoint -- 34 unique no fire plots  or SY < YearBurnedSecondMax 
q <- r3 |>
  filter(is.na(tFire))
# print q to see valid NA's -- in control (no burn history -34 plots)
# when sober, SampleYear < 2016 and no prior burn history
# when sober_chalk, SampleYear < 2008 and no prior burn history
# when dolan_chalk, SampleYear < 2008 and no prior burn history
# when chalk, SampleYear < 2008 and no prior burn history


# left join to add to p to track na's
rF <- r3 |> select(Plot, SampleYear, BurnScar, tFire) 
# 157/899 tFire are NA due to above reasons
purg <- left_join(plot, rF, by = c("Plot", "SampleYear", "BurnScar"))

#state factors
gg <- read.csv('./data/clean/bsplots-gg.csv') |> select(Plot, grtgroup)
soc <- read.csv('./data/clean/bsplots-soc.csv') |> select(Plot, b0, b10, b30)
prism <- read.csv('./data/clean/bsplots-prism.csv') |> select(Plot, precip, temp)
eco <- read.csv('./data/clean/bsplots-ndvi_albedo.csv') |> select(Plot, NDVI, albedo)

state <- gg |>
  left_join(soc, by = "Plot") |>
  left_join(prism, by = "Plot") |>
  left_join(eco, by = "Plot") |>
  rename(Albedo = albedo,
         Precipitation = precip,
         Temperature = temp)

# forgot redwood is 2, me is 1
type <- raw_plot |> select(PlotBS, ForestType)

# PRam, Fire, BurnScar, grtgroup, ForestType as factors
# Plot? SampleYear?
meta <- left_join(purg, state, by = "Plot") |>
  left_join(type, by = c("Plot" = "PlotBS")) |>
  mutate(across(c(PRam, Fire, BurnScar, grtgroup, ForestType), as.factor))

# meta is exported as load_plot() to bigsurrr

```



### Tree Database    
- Three datasets provided by the access database (tree1, tree2, tree3)    
- One provided by this [queried excel sheet](https://docs.google.com/spreadsheets/d/1HnFcAt-QIB9nOsEm94CMsk8XmoRtvoFf/edit?gid=1239576248#gid=1239576248) 

```{r}
# 116k total, 114k distinct, from google sheets
treeDB <- read_parquet("./data/arrow/tree_basic.parquet")
# 113k total and distinct
tree1 <- read_parquet("./data/arrow/tree_stems.parquet") |>
  select(StemSampleID, TreeSampleID, StemID, StemNumber, DBH, Status, DeadClass, LogDecayClass, SnagHeight,
         StumpHeight, StumpDiameter, StumpDecay, Prostrate, Canker)
# error mismatch of cols, tree2 only has TreeID and TreeNumber, but only way to get species
tree2 <- read_parquet("./data/arrow/tree_trees.parquet")
# can get other health columns merge on treeSampleID and treeID
tree3 <- read_parquet("./data/arrow/tree_samples.parquet")

# Need to quantify biomass, use allometric equations to convert DBH into biomass (kg C)
#allometric equations https://www.fs.usda.gov/nrs/pubs/jrnl/2014/nrs_2014_chojnacky_001.pdf
# equation is biomass(kg) = exp(b0 + b1 * log(DBH)) 
# parameters can be found on page 140 
eq <- function(b0,b1,DBH){
  exp(b0+b1*log(DBH))
}


# Only plan on studying Redwood, tanoak, bay laurel, ponderosa pine, douglas fir, madrone
tree_final <- treeDB |> 
  mutate(Species = str_replace(Species,"quag", "QUAG")) |>
  filter(Species %in% c("SESE","LIDE","UMCA","PIPO","PSME", "ARME", "QUAG")) |>
  mutate(Biomass = case_when(
    Species == "LIDE" ~ eq(-2.2198, 2.4410, DBH),
    Species == "SESE" ~ eq(-2.7765,2.4195, DBH),
    Species == "QUAG" ~ eq(-2.2198, 2.4410, DBH),
    Species == "UMCA" ~ eq(-2.2118, 2.4133, DBH),
    Species == "ARME" ~ eq(-2.2118, 2.4133, DBH),
    Species == "PSME" ~ eq(-2.4623, 2.4852, DBH),
    Species == "PIPO" ~ eq(-3.2007, 2.5339, DBH)
),
    Mortality = ifelse(Status  == "L", 0, 1)) |>
  mutate(across(c(Species, Status, Mortality), as.factor))

# this is exported to bigsurrr as load_tree()
```



### CWD Database   

- Two datasets provided by the access database (cwd, cwd_samples)   
- One dataset provided by Dr.Cobb (cwd_density)      

```{r}
# CWD Database 
cwd <- read_parquet("./data/arrow/cwd.parquet") |>
  rename(Plot = 2,
         Species = 3) |>
  select(CWD_ID,Plot,Species)

cwd_samples <- read_parquet("./data/arrow/cwd_samples.parquet") |>
  select(CWD_ID,SampleYear,DecayClass,Volume)

# merge on cwdid to add plot,species id's
# these id's are unique to stems
#only evaluating decayclass 1-5 due to density data gaps (only 1-5 studied) and
# QUAG, UMCA, SESE, LIDE -- other oaks?
cwd_final <- cwd_samples |>
  left_join(cwd, by="CWD_ID")|>
  filter(DecayClass <= 5 & Species %in% c('QUAG', 'UMCA', 'LIDE', 'SESE'))

# densities per species per decay class - experimental dataset provided by my professor
cwd_density <- read_parquet("./data/arrow/cwd_density.parquet") |>
  pivot_longer(cols=X1:X5, values_to="Density", names_to="DecayClass") |>
  rename(Species = Speices) |>
  mutate(DecayClass = str_remove(DecayClass, "X"))

cwd_meta <- left_join(cwd_final, cwd_density, by = c("Species", "DecayClass")) |>
  mutate(Mass = (Volume * (Density))) |>
  mutate(across(c(DecayClass, Species), as.factor))

# this is exported as load_cwd() to bigsurrr

# Only to follow repeatedly measured cwd stems - at least measured twice 
# Only need id, sampleyear, DecayClass,Plot,Species,Mass
# Need to create t column where t0 is the first time the CwdId was measured 
# need to create column named y which is log(Mt / M0) 
cwd_rep <- cwd_meta |>
  group_by(CWD_ID) |>
  filter(n() > 1) |>
  group_by(CWD_ID, Plot, DecayClass) |>
  mutate(
    tInit = min(SampleYear),
    t = SampleYear - tInit,
    Mass0 = Mass[t == 0],
    Y = log(Mass / Mass0)
  ) |>
  ungroup() 

# this is exported as load_cwd_repeated() in bigsurrr

```

### Meta Dataset    
- Only analyze LIDE, QUAG, SESE, UMCA in both tree and cwd pools    
  - Generate per ha CWD summaries, host summaries (QUAG, LIDE)     
  - Generate per ha dead host biomass   
- Remove plots with no CWD data   


```{r}
cwd_bio <- cwd_meta |>
  group_by(SampleYear, Plot) |>
  summarize(CWD_Ha_Biomass = sum(Mass) * 20)

cwd_host <- cwd_meta |>
  filter(Species %in% c('LIDE', 'QUAG')) |>
  group_by(SampleYear, Plot) |>
  summarize(CWD_Host_Biomass = sum(Mass) * 20) |>
  mutate(CWD_Host_Biomass = replace_na(CWD_Host_Biomass, 0))

cwd_sum <- left_join(cwd_bio, cwd_host, by = c("SampleYear", "Plot"))

tree_host <- tree_final |>
  filter(Species %in% c('LIDE', 'QUAG') & Status != "L") |>
  group_by(SampleYear, Plot) |>
  summarize(Snag_Host_Biomass = sum(Biomass) * 20) |>
  mutate(Snag_Host_Biomass = replace_na(Snag_Host_Biomass, 0))

frank <- left_join(tree_final, tree_host, by = c("Plot", "SampleYear")) |>
  left_join(cwd_sum, by = c("Plot", "SampleYear")) |>
  left_join(meta, by = c("Plot", "SampleYear")) |> 
  distinct()

#280 distinct plots with data
# for final df, filter out NA's
# 98,986 to 22, 497 datapoints
# 280 to 94 plots
f2 <- frank |>
  drop_na() |>
  distinct()


# this is load_meta() in bigsurrr
print(paste("Number of distinct plots:", n_distinct(f2$Plot)))
print(paste("Dimensions:", dim(f2)))
print("Sample Years:")
print(sort(unique(f2$SampleYear)))
print("Burn Scars:")
print(unique(f2$BurnScar))


```






