---
format: 
  html:
    toc: true 
editor: visual
execute:
  warning: false
  message: false
---
### Needed libraries   
```{r}
library(tidyverse)
library(broom)
library(gtsummary)
library(survival)
library(ggsurvfit)
library(lme4) 
library(lmerTest)
library(here)
library(knitr)
library(patchwork)
library(gt)

```    



### Recycled functions  
```{r}
read_me <- function(path){
  read_csv(here("data", path))
}
export_image <- function(img, path){
  ggsave(filename = here("exports", path), plot=img, width=12, height=8)
}

climate_theme <- function(color = "darkolivegreen3") {
  theme(
  plot.title = element_text(
      size = 16,
      family = "Times New Roman",
      color = "black", 
      margin = margin(b = 15)
    ),
    # add border 1)
    panel.border = element_rect(colour = "black", fill = NA, linetype = 1),
    # color background 2)
    panel.background = element_rect(fill = color),
    # modify grid 3)
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y =  element_blank(),
    panel.grid.minor.y = element_blank(),
    # modify text, axis and colour 4) and 5)
    axis.text = element_text(colour = "black",  family = "Times New Roman"),
    axis.title = element_text(colour = "black", family = "Times New Roman"),
    axis.ticks = element_line(colour = "black"),
    legend.position = "right"
  )
}
# darkolivegreen3
```


### CWD Biomass changes
```{r}
cwd_fire <- read_me("clean/cwd_mass_change_fire.csv")

burn_cwd_change <- function(burn, time1) {
  df <- cwd_fire |>
    filter(BurnScar == burn) |>
    mutate(Time = case_when(
      SampleYear <= time1  ~ "Pre",
      SampleYear >= time1 ~ "Post",
      TRUE ~ "NA")) 
  df_mod1 <- df |>
    filter(Time %in% c("Pre","Post")) |>
    group_by(Plot, BurnScar, Time) |>
    summarise(
      Biomass = sum(PlotCwdMass, na.rm = TRUE)) |>
    pivot_wider(names_from="Time", values_from="Biomass",names_expand = TRUE)
  df_mod2 <- df_mod1 |>
    filter(!is.na(Pre) & !is.na(Post))
  df_mod3 <- df_mod2 |>
    mutate(ChangeBiomass = Post - Pre) 
  return(df_mod3)
}

sites_cwd <- tibble(
  burn  = c("dolan", "sober",  "control", "dolan_chalk", "sober_chalk"),
  time1 = c(2020, 2016, 2016, 2009, 2009)
)

cwd_demo <- pmap_df(sites_cwd, burn_cwd_change)

post_fire_cwd_summary <- cwd_demo |>
  group_by(BurnScar) |>
  summarize( n = n(),
            MeanBiomassChange = mean(ChangeBiomass, na.rm=TRUE),
            MedianBiomassChange = median(ChangeBiomass, na.rm = TRUE),
            SdBiomassChange = sd(ChangeBiomass, na.rm=TRUE),
            MinBiomassChange = min(ChangeBiomass, na.rm=TRUE),
            MaxBiomassChange = max(ChangeBiomass, na.rm=TRUE)
) |>
  rename(`Burn Scar` = 1,
          `Mean Change in Cwd Biomass` = 3,
     `Median Change in Cwd Biomass` = 4,
      `SD Change in Cwd Biomass` = 5,
     `Min Change in Cwd Biomass` = 6,
     `Max Change in Cwd Biomass` = 7) 

kable(post_fire_cwd_summary)

cwd_uni <- cwd_demo |>
    ggplot(aes(x=BurnScar, y= ChangeBiomass, color = BurnScar)) +
    geom_boxplot(show.legend = FALSE) +
    geom_jitter(show.legend = FALSE) + 
    climate_theme() +
    labs(title= "Pre to Post Fire CWD Changes")

print(cwd_uni)
# anova testing
anov <- aov(ChangeBiomass ~ BurnScar, data = cwd_demo)
print(summary(anov))
print(TukeyHSD(anov))
norm <- cwd_demo |>
  ggplot(aes(x=ChangeBiomass)) +
  geom_histogram() +
  climate_theme() +
  labs(title = "Distribution of Response Var")
print(norm)


```

### CWD Decomposition   
-- Influence of species,disease,fire on decomposition rate of CWD.    
-- Linearized exponential decay model iterated across the species.    
-- Aim to estimate values of k and compare between plot conditions.   
**Questions I hope to answer**  
How do fire, disease and species impact decomposition rates?   
Is there a linear relationship between change in mass and time?     

-- This piggy backs off earlier research conducted by Dr.Cobb. I am utilizing a known equation for [leaf litter decay](https://cdnsciencepub.com/doi/10.1139/x06-012). Results are simple linear regression, just a lot of build up.    
-- Improvements: Split the dataset into burned/unburned, run individual models for each


**Size Assumptions**    
- Anything 3rd quartile or higher is large, 1-3 medium, 0-1 is small    


| Species | Large logs (Mg C) | Medium logs (Mg C) | Small logs (Mg C) |
|------------------|------------------|------------------|------------------|
| Redwood | > 0.46 | > 0.046 < 0.46 | < 0.046 |
|Tanoak | > 0.208 | > 0.0325 < 0.208 | < 0.0325 |
|Bay laurel | > 0.13 | > 0.027 < 0.13 | < 0.027 |

```{r}
cwd_decay <- read_me("./clean/cwd_decay.csv") |>
  mutate(SizeClass = case_when(
                               Mass <= 0.04 & Species == "Redwood" ~ "Small Logs",
                               Mass >= 0.04 & Mass <= 0.46 & Species == "Redwood"~ "Medium Logs",
                               Mass > 0.46 & Species == "Redwood" ~ "Large Logs",
                              Mass <= 0.0325 & Species == "Tanoak" ~ "Small Logs",
                               Mass >= 0.0325 & Mass <= 0.208 & Species == "Tanoak"~ "Medium Logs",
                               Mass > 0.208 & Species == "Tanoak" ~ "Large Logs",
                              Mass <= 0.027 & Species == "Bay laurel" ~ "Small Logs",
                               Mass >= 0.027 & Mass <= 0.13 & Species == "Bay laurel"~ "Medium Logs",
                               Mass > 0.13 & Species == "Bay laurel" ~ "Large Logs"))


decay <- function(species, table = TRUE){
  df <- cwd_decay |>
    filter(Species == species) 
  model <- nls(Y ~ -k*t, data=df, start=list(k=0.7))
  exp_model <- nls(Mass ~ a * exp(-k * t), start = list(a = 0.2, k=0.7),
              data = df)
  df$PredictedYLinear <- predict(model)
  summary <- summary(model)
  df$kLinear <- coef(model)[1]
  df$PredictedYExp <- predict(exp_model)
  summaryexp <- summary(exp_model)
  df$kExp <- coef(exp_model)[2]

  if (table){
    return(df)
  } else {
   new <- tibble(Equation = c("Linear", "Exponential"),
                 Species = c(species, species),
              n = rep(n_distinct(df$CwdId), 2), 
              RMSE = c(sqrt(mean(residuals(model) ^ 2)), (sqrt(mean(residuals(exp_model) ^ 2)))), 
              AIC = c(AIC(model), AIC(exp_model)),
              k = c(coef(model)[1], (coef(exp_model)[2]))
    )
    return(new)
  }
}




# apply model function to all species
species_vec <- unique(cwd_decay$Species)
results_summary <- map(.x = species_vec, .f = \(x) decay(x, table = FALSE)) |> bind_rows() 
results_raw <- map(.x = species_vec, .f = \(x) decay(x, table = TRUE)) |> bind_rows() 

results_raw |>
  ggplot(aes(x=t,y=Y, color = Species, linetype = Species)) +
  geom_point(shape=1, size=1.3) +
  geom_line(aes(x=t,y=PredictedYLinear, linetype = Species)) +
  facet_wrap(~SizeClass) + 
  labs(title="Linearized Exponential Decomposition Models for Studied Big Sur Trees", x="Time (t)", y="", subtitle="ln(Proportion of Mass Change in Year t)") +
 # geom_text(x = 3, y = 1, aes(label = paste("k = ", round(kLinear,4)))) +
  theme_classic()

results_raw |>
  ggplot(aes(x=t,y=Mass)) +
  geom_point(shape=1, size=1.3) +
  geom_line(aes(x=t,y=PredictedYExp), color="black", linetype="dashed") +
  facet_wrap(~Species) + 
  labs(title="Exponential Decomposition Models for Studied Big Sur Trees", x="Time (t)", y="", subtitle="Mass (Mg C ha-1)") +
  #geom_text(x = 3, y = 1, aes(label = paste("k = ", round(kExp,4)))) +
  theme_classic()




results_summary |>
  group_by(Equation) |>
  gt()



```


**Decomposition by size class**   
```{r}
df <- cwd_decay |>
  mutate(SizeClass = case_when(
                               Mass <= 0.04 ~ "Small Logs",
                               Mass >= 0.04 & Mass <= 0.36 ~ "Medium Logs",
                               Mass > 0.36 ~ "Large Logs"))

expo_decay_size <- function(size){
  df1 <- df |>
    filter(SizeClass == size) 
  model <- nls(Y ~ -k*t, data=df1, start=list(k=0.7))
  df1$PredictedY <- predict(model)
  summary <- summary(model)
  df1$k <- coef(model)[1]
  
  return(df1)
}

size_vec <- unique(df$SizeClass)
expo_results2 <- map_df(size_vec, expo_decay_size)


expo_results2 |>
  ggplot(aes(x=t,y=Y)) +
  geom_point(shape=1, size=1.3) +
  geom_line(aes(x=t,y=PredictedY), color="black", linetype="dashed") +
  facet_wrap(~SizeClass) + 
  labs(title="Linearized Exponential Decomposition Models for Studied Big Sur Trees", x="Time (t)", y="", subtitle="ln(Proportion of Mass Change in Year t)") +
  geom_text(x = 3, y = 1, aes(label = paste("k = ", round(k,4)))) +
  climate_theme()

#ggsave("./cwd_decay_size.png", p, width = 15, height = 12)

```

### CWD Mixed Modeling  

#### Plot Level CWD Mass    
Which factors influence plot level cwd mass? How do these factors influence cwd mass?   

```{r}
cwd_raw <- read_me("clean/cwd-mixed.csv")

cwd_simple <- function(df){
  df1 <- df |>
    mutate(across(c(PlotCwdMass, HostCwdMass, DeadHostBiomass, 
                  Elevation, Slope), scale),
    across(c(PRam, Fire, SampleYear), as.factor))
  tryCatch({
    m1 <- lmer(PlotCwdMass ~ Elevation + Slope + Fire + PRam +
                    (1|SampleYear) + (1|Plot),
                    data=df1
                     ) 
    summary <- summary(m1)
    df_summary <- as.data.frame(summary$coefficients) |>
      mutate(AIC = AIC(m1))
    df_summary$Term = rownames(df_summary)
    return(df_summary)
  }, error = function(e) {
    message("Error", e$message)
    return(NULL)
  })
}

cwd_host <- function(df){
  df1 <- df |>
    mutate(across(c(PlotCwdMass, HostCwdMass, DeadHostBiomass, 
                  Elevation, Slope), scale),
    across(c(PRam, Fire, SampleYear), as.factor))
  tryCatch({
    m1 <- lmer(PlotCwdMass ~ HostCwdMass + DeadHostBiomass + Elevation + Slope + Fire + PRam +
                    (1|SampleYear) + (1|Plot),
                    data=df1
                     ) 
    summary <- summary(m1)
    df_summary <- as.data.frame(summary$coefficients) |>
      mutate(AIC = AIC(m1))
    df_summary$Term = rownames(df_summary)
    return(df_summary)
  }, error = function(e) {
    message("Error", e$message)
    return(NULL)
  })
}

cwd_int <- function(df){
  df1 <- df |>
    mutate(across(c(PlotCwdMass, HostCwdMass, DeadHostBiomass, 
                  Elevation, Slope), scale),
    across(c(PRam, Fire, SampleYear), as.factor))
  tryCatch({
    m1 <- lmer(PlotCwdMass ~ Fire * PRam + HostCwdMass + DeadHostBiomass + Elevation + Slope +  
                    (1|SampleYear) + (1|Plot),
                    data=df1
                     ) 
    summary <- summary(m1)
    df_summary <- as.data.frame(summary$coefficients) |>
      mutate(AIC = AIC(m1))
    df_summary$Term = rownames(df_summary)
    return(df_summary)
  }, error = function(e) {
    message("Error", e$message)
    return(NULL)
  })
}


cwd_hostint <- function(df){
  df1 <- df |>
    mutate(across(c(PlotCwdMass, HostCwdMass, DeadHostBiomass, 
                  Elevation, Slope), scale),
    across(c(PRam, Fire, SampleYear), as.factor))
  tryCatch({
    m1 <- lmer(PlotCwdMass ~ HostCwdMass * DeadHostBiomass  * PRam + Fire + Elevation + Slope +
                    (1|SampleYear) + (1|Plot),
                    data=df1
                     ) 
    summary <- summary(m1)
    df_summary <- as.data.frame(summary$coefficients) |>
      mutate(AIC = AIC(m1))
    df_summary$Term = rownames(df_summary)
    return(df_summary)
  }, error = function(e) {
    message("Error", e$message)
    return(NULL)
  })
}

#cleaning function for streamlined results
clean_me <- function(df) {
  aic_val <- unique(df$AIC)
  df1 <- df |>
    rename(p = 5) |>
    mutate(Results = ifelse(
      p < 0.05,
      paste0("**", round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")**"),
      paste0(round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")")
    )) |>
    select(Term,Results) |>
    pivot_wider(names_from="Term", values_from="Results") |>
    mutate(AIC = round(aic_val, 2))
  
  return(df1)

}

sim <- clean_me(cwd_simple(cwd_raw))
host <- clean_me(cwd_host(cwd_raw))
int <- clean_me(cwd_int(cwd_raw))
host_int <- clean_me(cwd_hostint(cwd_raw))

model_comp <- bind_rows(
  sim |> mutate(Model = "Simple Model"),
  host |> mutate(Model = "Host Dead Biomass/Cwd Biomass Model"),
  int |> mutate(Model = "Fire/Disease Interaction Model"),
  host_int |> mutate(Model = "Host Interaction Model")
) |> 
  gt()

model_comp



```








