---
format: 
  html:
    toc: true
editor: visual
execute:
  warning: false
  message: false
---

### Needed libraries   
```{r}
library(tidyverse)
library(here)
library(readxl)
library(sf)
library(leaflet)
library(gt)
```    

```{r}
#speed up importing data
read_me <- function(path, status){
  read.csv(here("data",status, path))
}
# speed up export to data/clean
export_me <- function(df,path){
  write.csv(df, here("data","clean", path))
}

```

### Random sampling monte carlo simulations   

```{r}
bias <- read_me("bias.csv", "clean") 
ex <- bias |> select(Plot, Lon, Lat)
sf_obj <- st_as_sf(ex, coords = c("Lon", "Lat"), crs = 4326)
#st_write(sf_obj, "data/plots.shp")

z_score <- function(df,x){
  df1 <- df |>
    summarize(
      pop_mean <- mean({{x}}, na.rm=TRUE),
      pop_sd <- sd({{x}}, na.rm=TRUE)) |>
        rename(Mean = 1, Sd = 2) 
  df <- df |>
    mutate(ZScore = ({{x}} - df1$Mean)/ df1$Sd)
  return(df)
}
#sd = 2.311 mean=3.21 
# plotted bias histogram to verify Poisson distribution
bias_mod1 <- z_score(bias, TotalTimesInventoried)

# simulate poisson distribution 280 total, mean of 3.2
set.seed(331)
sim <- data.frame(Simulated = rpois(280,3.2)) 
sim_mod1 <- z_score(sim,Simulated)

# simulated zscore dataset is slightly right skewed, project dataset is heavily right skewed
# simulated zscore dataset has no values over ~3 while project dataset has values near 4
# radius is built on times inventoried
# color will be built on z-score
# project dataset, if z-score < -0.5 they get green
# if z-score > 3 they get red
# use of binning 3
bias_mod2 <- bias_mod1 |>
  mutate(Color = case_when(
                           ZScore < -0.5 ~ "Green",
                           ZScore > 2.75 ~ "Red",
                           TRUE ~ "Blue"
                           ))
#export_me(bias_mod2, "q7-map.csv")

# need this for .docx render
bias_hist <- bias_mod1 |>
  select(TotalTimesInventoried, ZScore) |>
  rename(Count = TotalTimesInventoried) |>
  mutate(Dataset = "Tree") 

sim_hist <- sim_mod1 |>
  rename(Count = Simulated) |>
  mutate(Dataset = "Simulated")
comp_hist <- bind_rows(bias_hist,sim_hist)


```


```{r}
map <- bias_mod2

leaflet(map) %>%
  addTiles() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  setView(lng = mean(map$Lon), lat = mean(map$Lat), zoom = 9) %>%
  addCircleMarkers(
    lng = ~Lon,
    lat = ~Lat,
    radius = ~Radius * 1.2,  # scale up for visibility
    label = ~Plot,
    popup = ~paste0(
      "<b>Plot:</b> ", Plot, "<br>",
      "<b>Times Inventoried:</b> ", TotalTimesInventoried, "<br>",
      "<b>Visit Frequency:</b> ", VisitFrequency
    ),
    color = ~Color,
    fillOpacity = 0.7
  ) |>
  addControl(
    html = "<b style='font-size:12px;'>Bias in Sampling Frequency</b>",
    position = "topright"
  ) 
  #|>
    #htmlwidgets::saveWidget("")

```

### State factors 

<a href=./state-factors.html target = "_blank">Extraction script</a>  

```{r}
#plot data
#CePlot 0=has ever had veg cov class <10%; 1= >10%; 2= >25%)

raw_plot <- read_me('raw_plotDB.csv', 'raw') |>
  slice(-1) |>
  select(1,2, 5, 27, 40, 50, 51, 63, 64) |>
  rename(PlotBS = 2,
         ForestType = 3,
         YearsBurned = 4,
         CePlot = 5,
         Trail = 6,
         Owner = 7,
         PlotAspect = 8,
         OverallAspect = 9) |>
  separate_wider_delim(YearsBurned, delim = ",", names_sep = "_", 
                       too_few = "align_start") |>
  pivot_longer(cols = YearsBurned_1:YearsBurned_6, names_to = "Year", 
               values_to = "YearBurned") |>
  mutate(YearBurned = str_trim(YearBurned)) |>
  drop_na(YearBurned)

type <- raw_plot |> select(PlotBS, ForestType, PlotAspect, OverallAspect)

plot <- read_me('plotDB.csv', 'raw')
topo <- plot  |> select(Plot, Elevation, Slope, Easting, Northing) |> distinct()
#state factors

gg <- read.csv('./data/bsplots-gg.csv') |> select(Plot, grtgroup)
soc <- read.csv('./data/bsplots-soc.csv') |> select(Plot, b0, b10, b30)
prism <- read.csv('./data/bsplots-prism.csv') |> select(Plot, precip, temp)
eco <- read.csv('./data/bsplots-ndvi_albedo.csv') |> select(Plot, NDVI, albedo)

#redwood is 2, me is 1
state <- gg |>
  left_join(soc, by = "Plot") |>
  left_join(prism, by = "Plot") |>
  left_join(eco, by = "Plot") |>
  left_join(type, by = c("Plot" = "PlotBS")) |>
  left_join(topo, by = "Plot") |>
  mutate(across(c(PlotAspect, OverallAspect), as.numeric)) |>
  mutate(ForestType = case_when(
                                ForestType == "1" ~ 0,
                                TRUE ~ 1)) |>
  mutate(across(c(grtgroup, ForestType), as.factor)) 


```

**Which state factors drive forest type?**    

$$
\begin{multline}
\text{Forest Type} \sim \text{Soil Great Group} + \text{SOC 0cm} + \text{SOC (< 10cm)} + \text{SOC (< 30cm)} + \\
\text{Precipitation} + \text{Temperature} + \text{NDVI} + \text{Albedo} +
\text{Plot Aspect} + \text{Elevation} + \text{Slope} + \text{Easting} + \text{Northing}
\end{multline}
$$


### Logistic Regression    
```{r}
library(car)
library(pROC)

state_model1 <- glm(ForestType ~ grtgroup + b0 + b10 + b30 + precip + temp +
                    NDVI + albedo + PlotAspect + Elevation + Slope + Easting + 
                    Northing, data = state, family = "binomial") 

log_vis <- summary(state_model1) 
print(log_vis)
print(vif(state_model1))

log_pred <- predict(state_model1, type = "response")

roc_obj <- roc(state$ForestType, log_pred, quiet = TRUE)
auc_val <- as.numeric(auc(roc_obj))
print(paste("ROC: ", round(auc_val, 3)))

plot(roc_obj, main = "ROC:",
     col = "#9370DB", lwd = 3, legacy.axes = TRUE)

```

### Random forest classification model    

```{r}
library(ranger)
library(vip)
set.seed(200)

train_index <- sample(1:nrow(state), 0.8 * nrow(state))
train <- state[train_index, ]
test <- state[-train_index, ]


rf_model <- ranger(ForestType ~., data = train, 
                   num.trees = 100, 
                    importance = "permutation"
)

prediction <- predict(rf_model, data = test)$predictions

rmse_rf <- sqrt(rf_model$prediction.error) 

accuracy <- mean(prediction == test$ForestType)
print(paste("Accuracy:", accuracy))


p1 <- vip::vip(rf_model, bar = FALSE)
print(p1)

roc_rf <- roc(test$ForestType, as.numeric(prediction), quiet = TRUE)
auc_rf <- as.numeric(auc(roc_rf))
print(paste("ROC: ", round(auc_rf, 3)))

plot(roc_rf, main = "ROC:",
     col = "#9370DB", lwd = 3, legacy.axes = TRUE)



```

**How do the logistic and Random forest models compare?**   
- Utilizing auc  

```{r}
comps <- tibble(
                   Model = c("Logistic Regression", "Random Forest"),
                             AUC = c(auc_val, auc_rf)) |>
gt()

comps
```


### Time since fire calculations   

```{r}
plotDB <- left_join(raw_plot, plot, by = c("PlotBS" = "Plot"), relationship = "many-to-many") |>
  separate_wider_delim(YearBurned, delim = " ", names_sep = "_",
                       too_few = "align_start") |>
  pivot_longer(cols = YearBurned_1:YearBurned_2, names_to = "trans", values_to = "YearBurned") |>
  drop_na(YearBurned, BurnScar) |>
  select(!c(Year, trans)) |>
  mutate(YearBurned = case_when(
         YearBurned %in% c("none", "") ~ "0",
         TRUE ~ YearBurned
       )) |>
  mutate(across(c(Plot, ForestType, PlotAspect, OverallAspect, YearBurned), as.numeric)) 


plot_t <- plotDB |>
  select(PlotBS, SampleYear, YearBurned, BurnScar) |>
  distinct() |>
  group_by(PlotBS, SampleYear, BurnScar) |>
  summarize(YearBurnedThirdMax = nth(sort(YearBurned, decreasing = TRUE), 3),
            YearBurnedSecondMax = nth(sort(YearBurned, decreasing = TRUE), 2),
            YearBurnedMax = max(YearBurned))

## time since fire calcs
recent <- plot_t |>
  mutate(tFire = case_when(
                           BurnScar == "dolan" & SampleYear >= 2020 ~ SampleYear - 2020,
                           BurnScar == "dolan" & SampleYear < 2020 ~ SampleYear - YearBurnedMax,
                           BurnScar == "dolan_chalk" & SampleYear >= 2020 ~ SampleYear - 2020,
                           BurnScar == "dolan_chalk" & SampleYear < 2020 & SampleYear >= 2008 ~ SampleYear - 2008,
                          BurnScar == "dolan_chalk" & SampleYear < 2008 ~ SampleYear - YearBurnedSecondMax,
                           BurnScar == "sober" & SampleYear >= 2016 ~ SampleYear - 2016,
                           BurnScar == "sober" & SampleYear < 2016 ~ SampleYear - YearBurnedSecondMax,
                           BurnScar == "sober_chalk" & SampleYear >= 2016 ~ SampleYear - 2016,
                           BurnScar == "sober_chalk" & SampleYear < 2016 & SampleYear >= 2008 ~ SampleYear - 2008,
                            BurnScar == "sober_chalk" & SampleYear < 2008 ~ SampleYear - YearBurnedThirdMax,
                            BurnScar == "chalk" & SampleYear >= 2008 ~ SampleYear - 2008,
                          BurnScar == "chalk" & SampleYear < 2008 ~ SampleYear - YearBurnedSecondMax,
                          BurnScar == "control" ~ SampleYear - YearBurnedMax)) |>
    select(PlotBS, BurnScar, SampleYear, tFire, YearBurnedMax, YearBurnedSecondMax, YearBurnedThirdMax) 

# 56 unique plots with only one burn date, unburnt before? 
na <- recent |>
  filter(is.na(tFire)) 

#4 -t values - control group burn years in 2013, entry error?
plotDB2 <- recent |>
  filter(!is.na(tFire) &
         tFire > 0) |> 
  select(PlotBS, SampleYear, tFire)

plot_meta <- left_join(plotDB, plotDB2, by = c("PlotBS", "SampleYear"))

```

## Principle Component Analysis     

**Objective**   
- Better understand variability within each dataset  
- Add groups beyond Plot ID to test as random factors 


**Plot Dataset**   

```{r}
#| fig-width: 18
#| fig-height: 8


library(corrr)
make_titles <- function(x){
  # expects a character or factor as an input
  stopifnot(is.character(x) | is.factor(x))
  
  # remove the units at the end of the name
  str_remove(x, pattern = "(g|mm)$") |> 
  # replace all _ with spaces
  str_replace_all(pattern = "_", replacement = " ") |> 
  # remove extra whitespace on left 
  str_trim(side = "both") |> 
  # convert the first letter of each word to upper case
  str_to_title()
}

cor_matrix <- function(df, title) {
  df |>
    select(where(is.numeric)) |>
    correlate(method = "pearson",
            use = "pairwise.complete.obs") |>
    rename(term1 = term) |> 
    pivot_longer(cols = -term1, 
               names_to = "term2", 
               values_to = "correlation") |>
    mutate(
      across(.cols = c(term1, term2), 
           .fns = ~ make_titles(.x)
           )
    ) |> 
    ggplot(mapping = aes(x = term1,
                       y = term2, 
                       fill = correlation)) +
      geom_tile() +
      labs(x = "", 
       y = "", 
       title = paste("Correlation Matrix of:", title), 
       fill = "Correlation"
       ) +
      theme_bw() +
      scale_fill_viridis_c(option = "B")

}

cor_matrix(plot_meta, "Plot Qualitative Variables")

```



```{r}
#| fig-width: 18
#| fig-height: 10

# only numeric cols, will auto scale, center and return transformed data
pca_me <- function(df) {
  df1 <- df |>
    ungroup() |>
    select(!CePlot) |>
    select(where(is.numeric)) |>
    drop_na() 
  pca <- prcomp(df1, scale. = TRUE, center = TRUE, retx = TRUE)
  
  pca_var <- pca$sdev^2
  propve <- pca_var / sum(pca_var)
  print(summary(pca))

  par(mfrow = c(1,2))
  plot(propve, xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     ylim = c(0, 1), type = "b", main = "Scree Plot")
 
  biplot(pca, xlabs=rep("", nrow(df1))) 
  
}


pca_me(plot_meta)


```


**Tree Dataset**

```{r}
#| fig-width: 18
#| fig-height: 10


tree <- read.csv('./data/clean/tree_table_FINAL.csv') |>
  rename(PlotBS = BS_plot,
         StemId = StemID,
         BasalArea = BasalArea.Plot) |>
  select(PlotBS, SampleYear, Species, StemId, GenetId, RametId, HealthClass, Status,
         DBH, BasalArea, RametFail, GenetFail) |>
  left_join(plot_meta, by = c("PlotBS", "SampleYear"), relationship = "many-to-many")

# Need to quantify biomass, use allometric equations to convert DBH into biomass (kg C)
#allometric equations https://www.fs.usda.gov/nrs/pubs/jrnl/2014/nrs_2014_chojnacky_001.pdf
# equation is biomass(kg) = exp(b0 + b1 * log(DBH)) 
eq <- function(b0,b1,DBH){
  exp(b0+b1*log(DBH))
}

tree2 <- tree |>
  mutate(Species = str_replace(Species,"quag", "QUAG")) |>
  filter(Species %in% c("SESE","LIDE","UMCA","PIPO","PSME", "ARME", "QUAG")) |>
  mutate(Biomass = case_when(
    Species == "LIDE" ~ eq(-2.2198, 2.4410, DBH),
    Species == "SESE" ~ eq(-2.7765,2.4195, DBH),
    Species == "QUAG" ~ eq(-2.2198, 2.4410, DBH),
    Species == "UMCA" ~ eq(-2.2118, 2.4133, DBH),
    Species == "ARME" ~ eq(-2.2118, 2.4133, DBH),
    Species == "PSME" ~ eq(-2.4623, 2.4852, DBH),
    Species == "PIPO" ~ eq(-3.2007, 2.5339, DBH)
))

cor_matrix(tree2, "Tree Qualitative Variables")

```

```{r}
#| fig-width: 18
#| fig-height: 10
pca_me(tree2)
```

**Living Biomass Dataset**

```{r}
#| fig-width: 18
#| fig-height: 10

live <- tree2 |>
  filter(Status == "L") |>
  group_by(PlotBS, SampleYear, Species) |>
  summarize(StemDensity = n(),
            BiomassHa = (sum(Biomass, na.rm = TRUE) * 20)) |>
  left_join(plot_meta, by = c("PlotBS", "SampleYear"), relationship = "many-to-many")


cor_matrix(live, "Living Biomass Qualitative Variables")

```

```{r}
#| fig-width: 18
#| fig-height: 10

pca_me(live)

```

**Standing Dead Biomass**

```{r}
#| fig-width: 18
#| fig-height: 10

dead <- tree2 |>
  filter(!Status == "L") |>
  group_by(PlotBS, SampleYear, Species) |>
  summarize(StemDensity = n(),
            StandingDeadBiomassHa = (sum(Biomass, na.rm = TRUE) * 20)) |>
  left_join(plot_meta, by = c("PlotBS", "SampleYear"), relationship = "many-to-many")


cor_matrix(dead, "Standing Dead Biomass Qualitative Variables")

```

```{r}
#| fig-width: 18
#| fig-height: 10

pca_me(dead)

```




**Coarse woody debris biomass**


```{r}
#| fig-width: 18
#| fig-height: 10

cwd <- read.csv('./data/raw/cwdDB.csv') |>
  group_by(Plot, SampleYear, Species) |>
  summarize(StemDensity = n(),
            CwdBiomassHa = (sum(MassPlot, na.rm = TRUE) * 20)) |>
  left_join(plot_meta, by = c("Plot" = "PlotBS", "SampleYear"), relationship = "many-to-many")


cor_matrix(cwd, "CWD Biomass Qualitative Variables")

```

```{r}
#| fig-width: 18
#| fig-height: 10

pca_me(cwd)

```








