---
format: 
  html:
    toc: true 
editor: visual
execute:
  warning: false
  message: false
---
### Needed libraries   
```{r}
library(tidyverse)
library(broom)
library(gtsummary)
library(survival)
library(ggsurvfit)
library(lme4) 
library(lmerTest)
library(here)
library(knitr)
library(patchwork)
library(gt)

```    

### Recycled functions  
```{r}
read_me <- function(path){
  read_csv(here("data", path))
}
export_image <- function(img, path){
  ggsave(filename = here("exports", path), plot=img, width=12, height=8)
}

climate_theme <- function(color = "slategray1") {
  theme(
  plot.title = element_text(
      size = 16,
      family = "Times New Roman",
      color = "black", 
      margin = margin(b = 15)
    ),
    # add border 1)
    panel.border = element_rect(colour = "black", fill = NA, linetype = 1),
    # color background 2)
    panel.background = element_rect(fill = color),
    # modify grid 3)
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y =  element_blank(),
    panel.grid.minor.y = element_blank(),
    # modify text, axis and colour 4) and 5)
    axis.text = element_text(colour = "black",  family = "Times New Roman"),
    axis.title = element_text(colour = "black", family = "Times New Roman"),
    axis.ticks = element_line(colour = "black"),
    legend.position = "right"
  )
}
# darkolivegreen3

```

### Logistic regression of mortality status  
-- Multiple logistic regression iterated across species   
    - Attempting to predict drivers of mortality given binary health status.  
- Improvements: Use mixed effects/survival analysis instead of logistic regression.   
**Questions I hope to answer**  
Which features, topographic or epidemiological, are driving mortality in Redwood forests?  
Is there a relationship between the explanatory variables and mortality?    

-- P-values for fixed variables below the 0.05 threshold will allow us to reject our null hypothesis and indicate a relationship between the explanatory variables and mortality.

```{r}
mortality <-  read_me("./clean/mortality_mixedmodel.csv")
species_mortality <- unique(mortality$Species)

mortality_model <- function(species){
  df1 <- mortality |>
    filter(Species == {{species}})
  logistic_model <- glm(Mortality ~ PlotCwdMass + HostCwdMass + Elevation + Slope + Fire + PRam,
                    data=df1, family = binomial
                     )
  summary <- summary(logistic_model)
  
  df_summary <- as.data.frame(summary$coefficients) |>
     mutate(Species = species)
  df_summary$Term = rownames(df_summary)
  return(df_summary)
}

mortality_results <- map_df(species_mortality, mortality_model)

mortality_results2 <- mortality_results |>
  rename(p = 4) |>
  mutate(Results = ifelse(
    p < 0.05,
    paste0("**", round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")**"),
    paste0(round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")")
  )) |>
  select(Species,Term,Results) |>
  pivot_wider(names_from="Term", values_from="Results") |>
  select(!`(Intercept)`)

mortality_results3 <- mortality_results2 |>
  rename(Disease = PRam,
        `Plot Cwd Mass` = 2,
        `Host Cwd Mass` = 3)

kable(mortality_results3)

```

## Tracking stems through time   

### Survival Analysis

#### 1. Need to load and transform data into survival tables  
  - Create t and set t0 = 2006

```{r}
tree <- read_me("./clean/tree_table_FINAL.csv")

tree1 <- tree |>
  filter(Species %in% c("SESE", "LIDE","UMCA")) |> 
  rename(Disease = p_ram.presence,
         Fire = fire_presence) |>
  filter(!is.na(Species)) |>
  mutate(t = SampleYear - 2006,
         Species = case_when(
                             Species == "SESE" ~ "Redwood",
                             Species == "LIDE" ~ "Tanoak",
                             TRUE ~ "Bay laurel")) |>
  select(BS_plot, Species, t,RametFail,Disease,Fire, DBH)

```

#### [Kaplian - Meier](https://www.emilyzabor.com/survival-analysis-in-r.html#part-4-advanced-topics)  

```{r}
s1 <- survfit2(Surv(t,RametFail) ~ Species, data=tree1)

s1 |>
  ggsurvfit() +
  facet_wrap(~strata) +
  theme_classic() +
  labs(title="Kaplan-Meier Survival Function of Big Sur Species", x="Time")

#export_image(p1, "surv.png") 

summary(s1)
survdiff(Surv(t, RametFail) ~ Species, data = tree1)

``` 






