---
format: 
  html:
    toc: true 
editor: visual
execute:
  warning: false
  message: false
---
### Needed libraries   
```{r}
library(tidyverse)
library(broom)
library(gtsummary)
library(survival)
library(ggsurvfit)
library(lme4) 
library(lmerTest)
library(here)
library(knitr)
library(patchwork)
library(gt)

```    



### Recycled functions  
```{r}
read_me <- function(path){
  read_csv(here("data", path))
}
export_image <- function(img, path){
  ggsave(filename = here("exports", path), plot=img, width=12, height=8)
}

climate_theme <- function(color = "slategray1") {
  theme(
  plot.title = element_text(
      size = 16,
      family = "Times New Roman",
      color = "black", 
      margin = margin(b = 15)
    ),
    # add border 1)
    panel.border = element_rect(colour = "black", fill = NA, linetype = 1),
    # color background 2)
    panel.background = element_rect(fill = color),
    # modify grid 3)
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y =  element_blank(),
    panel.grid.minor.y = element_blank(),
    # modify text, axis and colour 4) and 5)
    axis.text = element_text(colour = "black",  family = "Times New Roman"),
    axis.title = element_text(colour = "black", family = "Times New Roman"),
    axis.ticks = element_line(colour = "black"),
    legend.position = "right"
  )
}
# darkolivegreen3
```


### CWD Biomass changes
```{r}

cwd_fire <- read_me("./clean/cwd_mass_change_fire.csv")

burn_cwd_change <- function(burn, time1) {
  df <- cwd_fire |>
    filter(BurnScar == burn) |>
    mutate(Time = case_when(
      SampleYear <= time1  ~ "Pre",
      SampleYear >= time1 ~ "Post",
      TRUE ~ "NA")) 
  df_mod1 <- df |>
    filter(Time %in% c("Pre","Post")) |>
    group_by(Plot, BurnScar, Time) |>
    summarise(
      Biomass = sum(PlotCwdMass, na.rm = TRUE)) |>
    pivot_wider(names_from="Time", values_from="Biomass",names_expand = TRUE)
  df_mod2 <- df_mod1 |>
    filter(!is.na(Pre) & !is.na(Post))
  df_mod3 <- df_mod2 |>
    mutate(ChangeBiomass = Post - Pre) 
  return(df_mod3)
}

sites_cwd <- tibble(
  burn  = c("dolan", "sober",  "control", "dolan_chalk", "sober_chalk"),
  time1 = c(2020, 2016, 2016, 2009, 2009)
)

cwd_demo <- pmap_df(sites_cwd, burn_cwd_change)

post_fire_cwd_summary <- cwd_demo |>
  group_by(BurnScar) |>
  summarize( n = n(),
            MeanBiomassChange = mean(ChangeBiomass, na.rm=TRUE),
            MedianBiomassChange = median(ChangeBiomass, na.rm = TRUE),
            SdBiomassChange = sd(ChangeBiomass, na.rm=TRUE),
            MinBiomassChange = min(ChangeBiomass, na.rm=TRUE),
            MaxBiomassChange = max(ChangeBiomass, na.rm=TRUE)
) |>
  rename(`Burn Scar` = 1,
          `Mean Change in Cwd Biomass` = 3,
     `Median Change in Cwd Biomass` = 4,
      `SD Change in Cwd Biomass` = 5,
     `Min Change in Cwd Biomass` = 6,
     `Max Change in Cwd Biomass` = 7) 

kable(post_fire_cwd_summary)

cwd_uni <- cwd_demo |>
    ggplot(aes(x=BurnScar, y= ChangeBiomass, color = BurnScar)) +
    geom_boxplot(show.legend = FALSE) +
    geom_jitter(show.legend = FALSE) + 
    climate_theme() +
    labs(title= "Pre to Post Fire CWD Changes")

print(cwd_uni)
# anova testing
anov <- aov(ChangeBiomass ~ BurnScar, data = cwd_demo)
print(summary(anov))
print(TukeyHSD(anov))
norm <- cwd_demo |>
  ggplot(aes(x=ChangeBiomass)) +
  geom_histogram() +
  climate_theme() +
  labs(title = "Distribution of Response Var")
print(norm)


```

### CWD Decomposition   
-- Influence of species,disease,fire on decomposition rate of CWD.    
-- Linearized exponential decay model iterated across the species.    
-- Aim to estimate values of k and compare between plot conditions.   
**Questions I hope to answer**  
How do fire, disease and species impact decomposition rates?   
Is there a linear relationship between change in mass and time?     

-- This piggy backs off earlier research conducted by Dr.Cobb. I am utilizing a known equation for [leaf litter decay](https://cdnsciencepub.com/doi/10.1139/x06-012). Results are simple linear regression, just a lot of build up.    
-- Improvements: Split the dataset into burned/unburned, run individual models for each


```{r}
cwd_decay <- read_me("./clean/cwd_decay.csv")
expo_decay <- function(species){
  df <- cwd_decay |>
    filter(Species == species) 
  model <- nls(Y ~ -k*t, data=df, start=list(k=0.7))
  df$PredictedY <- predict(model)
  summary <- summary(model)
  
  return(df)
}
# apply model function to all species
species_vec <- unique(cwd_decay$Species)
expo_results <- map_df(species_vec, expo_decay)
# need to prepare for visuals
expo_results2 <- expo_results |>
  mutate(Fire = as.factor(Fire)) |>
  drop_na() |>
  mutate(Burn = ifelse(Fire == 0, "Unburned", "Burned"))


expo_results2 |>
  ggplot(aes(x=t,y=Y, color=Burn)) +
  geom_point(shape=1, size=1.3) +
  geom_line(aes(x=t,y=PredictedY), color="red", linetype="dashed") +
  facet_wrap(~Species,scales = "free") + 
  labs(title="Linearized Exponential Decomposition Models for Studied Big Sur Trees", x="Time (t)", y="", subtitle="ln(Proportion of Mass Change in Year t)",
  color="Fire Status") +
  theme_classic()

```

### CWD Mixed Modeling  

```{r}


```








