---
format: 
  html:
    code-tools: true
    embed-resources: true
    toc: true 
    theme: flatly
    page-layout: full
editor: visual
execute:
  warning: false
  message: false
---

## Objectives: quantifying fire/disease impacts   
- Quantify biomass, stem density changes pre to post fire. Compare across burn scar, species, and status.   
  - Used mixed effects models to understand underlying relationships.   
- Track stems through time to quantify rates of mortality, snag fall and CWD decomposition. 
  - Use survival analysis (mortality), ?? (snag fall), regression (CWD decomp.)    

### Needed libraries   
```{r}
library(tidyverse)
library(broom)
library(gtsummary)
library(survival)
library(ggsurvfit)
library(lme4) 
library(lmerTest)
library(here)
library(knitr)
library(patchwork)
library(gt)

```    

### Recycled functions  
```{r}
read_me <- function(path){
  read_csv(here("data", path))
}
export_image <- function(img, path){
  ggsave(filename = here("exports", path), plot=img, width=12, height=8)
}

climate_theme <- function(color = "slategray1") {
  theme(
  plot.title = element_text(
      size = 16,
      family = "Times New Roman",
      color = "black", 
      margin = margin(b = 15)
    ),
    # add border 1)
    panel.border = element_rect(colour = "black", fill = NA, linetype = 1),
    # color background 2)
    panel.background = element_rect(fill = color),
    # modify grid 3)
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y =  element_blank(),
    panel.grid.minor.y = element_blank(),
    # modify text, axis and colour 4) and 5)
    axis.text = element_text(colour = "black",  family = "Times New Roman"),
    axis.title = element_text(colour = "black", family = "Times New Roman"),
    axis.ticks = element_line(colour = "black"),
    legend.position = "right"
  )
}
# darkolivegreen3

```

## Univariate statistics of fire and disease impacts 

### Living Biomass changes   

```{r}
#| fig-width: 14
#| fig-height: 16


biomass_changes <- read_me("./clean/post_fire_summary.csv")
burn_change <- function(burn, time1) {
  df <- biomass_changes |>
    filter(BurnScar == burn) |>
    select(!c(Elevation,Slope,Fire)) |>
    mutate(Pre = case_when(
      SampleYear <= time1 ~ "Pre",
      SampleYear >=  time1 ~ "Post",
      TRUE ~ "NA")) 
  df_mod1 <- df |>
    filter(Pre %in% c("Pre","Post")) |>
    group_by(Plot, Species, BurnScar, PRam, Pre) |>
    summarise(
      TotalStem   = sum(TotalStem, na.rm = TRUE),
      LiveBiomass = sum(LiveBiomass, na.rm = TRUE),
      .groups = "drop"
    ) |>
    pivot_wider(names_from="Pre", values_from=c("TotalStem","LiveBiomass"))
  # only want to evaluate data with both pre and post 
  df_mod2 <- na.omit(df_mod1)
  df_mod3 <- df_mod2 |>
    mutate(ChangeStem = TotalStem_Post - TotalStem_Pre,
      ChangeBiomass = LiveBiomass_Post - LiveBiomass_Pre) |>
    select(c(1,2,3,4,9,10))
  return(df_mod3)
}
# apply summary function to dataset we want
sites_tree <- tibble(
  burn  = c("dolan", "sober", "chalk", "control", "dolan_chalk", "sober_chalk"),
  time1 = c(2020, 2016, 2007, 2016, 2007, 2007)
)
demo_tree <- pmap_df(sites_tree, burn_change)

post_fire_summary_biomass <- demo_tree |>
  group_by(Species, BurnScar, PRam) |>
  summarize( n = n(),
            MeanStemChange = mean(ChangeStem, na.rm=TRUE),
            MeanBiomassChange = mean(ChangeBiomass, na.rm=TRUE),
            MedianStemChange    = median(ChangeStem, na.rm = TRUE),
            MedianBiomassChange = median(ChangeBiomass, na.rm = TRUE),
            SdStemChange = sd(ChangeStem, na.rm=TRUE),
            SdBiomassChange = sd(ChangeBiomass, na.rm=TRUE)
) |>
  rename(
         `Mean Change in Stem Density` = 3,
     `Mean Change in Biomass` = 4,
     `Median Change in Stem Density` = 5,
     `Median Change in Biomass` = 6,
  `SD Change in Stem Density` = 7,
  `SD Change in Biomass` = 8)


kable(post_fire_summary_biomass)


den <- demo_tree |>
    ggplot(aes(x=ChangeStem, y= ChangeBiomass, color = Species, linetype = as.factor(PRam))) +
    geom_point() +
    geom_smooth(method = "lm", se =FALSE) +
    climate_theme() 

dis <- demo_tree |>
    ggplot(aes(x=BurnScar, y= ChangeBiomass, color = BurnScar, fill = as.factor(PRam))) +
    geom_boxplot() +
    geom_jitter(shape = 21, 
                position = position_jitterdodge(0.8)) + 
    climate_theme() +
    scale_color_viridis_d()

dis2 <- demo_tree |>
    ggplot(aes(x=BurnScar, y= ChangeStem, color = BurnScar, fill = as.factor(PRam))) +
    geom_boxplot() +
    geom_jitter(shape = 21, 
                position = position_jitterdodge(0.8)) + 
    climate_theme()
dis3 <- demo_tree  |>
    ggplot(aes(x=Species, y= ChangeBiomass, color = Species, fill = as.factor(PRam))) +
    geom_boxplot() +
    geom_jitter(shape = 21, 
                position = position_jitterdodge(0.8)) + 
    climate_theme()

dis4 <- demo_tree |>
    ggplot(aes(x=Species, y= ChangeStem, color = Species, fill = as.factor(PRam))) +
    geom_boxplot() +
    geom_jitter(shape = 21, 
                position = position_jitterdodge(0.8)) + 
    climate_theme()


(dis + dis2) / (dis3 + dis4) / den +
  plot_annotation(title = "Pre to Post Fire Living Tree Summary") +
plot_layout(heights = c(1, 1, 2)) 


```


### CWD Biomass changes
```{r}

cwd_fire <- read_me("./clean/cwd_mass_change_fire.csv")

burn_cwd_change <- function(burn, time1) {
  df <- cwd_fire |>
    filter(BurnScar == burn) |>
    mutate(Time = case_when(
      SampleYear <= time1  ~ "Pre",
      SampleYear >= time1 ~ "Post",
      TRUE ~ "NA")) 
  df_mod1 <- df |>
    filter(Time %in% c("Pre","Post")) |>
    group_by(Plot, BurnScar, Time) |>
    summarise(
      Biomass = sum(PlotCwdMass, na.rm = TRUE)) |>
    pivot_wider(names_from="Time", values_from="Biomass",names_expand = TRUE)
  df_mod2 <- df_mod1 |>
    filter(!is.na(Pre) & !is.na(Post))
  df_mod3 <- df_mod2 |>
    mutate(ChangeBiomass = Post - Pre) 
  return(df_mod3)
}

sites_cwd <- tibble(
  burn  = c("dolan", "sober",  "control", "dolan_chalk", "sober_chalk"),
  time1 = c(2020, 2016, 2016, 2009, 2009)
)

cwd_demo <- pmap_df(sites_cwd, burn_cwd_change)

post_fire_cwd_summary <- cwd_demo |>
  group_by(BurnScar) |>
  summarize( n = n(),
            MeanBiomassChange = mean(ChangeBiomass, na.rm=TRUE),
            MedianBiomassChange = median(ChangeBiomass, na.rm = TRUE),
            SdBiomassChange = sd(ChangeBiomass, na.rm=TRUE),
            MinBiomassChange = min(ChangeBiomass, na.rm=TRUE),
            MaxBiomassChange = max(ChangeBiomass, na.rm=TRUE)
) |>
  rename(`Burn Scar` = 1,
          `Mean Change in Cwd Biomass` = 3,
     `Median Change in Cwd Biomass` = 4,
      `SD Change in Cwd Biomass` = 5,
     `Min Change in Cwd Biomass` = 6,
     `Max Change in Cwd Biomass` = 7) 

kable(post_fire_cwd_summary)

cwd_uni <- cwd_demo |>
    ggplot(aes(x=BurnScar, y= ChangeBiomass, color = BurnScar)) +
    geom_boxplot(show.legend = FALSE) +
    geom_jitter(show.legend = FALSE) + 
    climate_theme() +
    labs(title= "Pre to Post Fire CWD Changes")

print(cwd_uni)
# anova testing
anov <- aov(ChangeBiomass ~ BurnScar, data = cwd_demo)
print(summary(anov))
print(TukeyHSD(anov))
norm <- cwd_demo |>
  ggplot(aes(x=ChangeBiomass)) +
  geom_histogram() +
  climate_theme() +
  labs(title = "Distribution of Response Var")
print(norm)


```



## Underlying drivers of biomass and stem density changes   

### Mixed effects models   
-- Multiple regression with mixed effects models iterated across species     
    - Attempt to understand drivers of biomass changes. Is disease and fire interacting to reduce biomass? Is it even across species.       
- Improvements: Run multiple models per species with different explanatory variables, take weighted AIC and weighted effects.  
**Questions I hope to answer**  
Which features, topographic or epidemiological, are driving biomass changes in Redwood forests?  
Is there a linear relationship between the fixed explanatory variables and biomass?  

-- P-values for fixed variables below the 0.05 threshold will allow us to reject our null hypothesis and indicate a linear relationship between a fixed explanatory variable and biomass.



#### Plot biomass changes 

- Scaling values standardizes units? Worth doing? 

```{r}
biomass <-  read_me("./clean/livebiomass_mixedmodel.csv")

species_df <- unique(biomass$Species)

big_biomass_model <- function(species){
  df1 <- biomass |>
    filter(Species == {{species}}) |>
    mutate(across(c(PlotCwdMass, HostCwdMass, DeadHostBiomass, 
                  Elevation, Slope,LiveBiomass), scale))
  tryCatch({
    m1 <- lmer(LiveBiomass ~ PlotCwdMass + HostCwdMass + DeadHostBiomass + Elevation + Slope + Fire + PRam +
                    (1|SampleYear) + (1|Plot),
                    data=df1
                     ) 
    summary <- summary(m1)
    df_summary <- as.data.frame(summary$coefficients) |>
      mutate(Species = species,
              AIC = AIC(m1))
    df_summary$Term = rownames(df_summary)
    return(df_summary)
  }, error = function(e) {
    message("Error", e$message)
    return(NULL)
  })
}

interaction_model <- function(species){
  df1 <- biomass |>
    filter(Species == {{species}}) |>
    mutate(across(c(PlotCwdMass, HostCwdMass, DeadHostBiomass, 
                  Elevation, Slope,LiveBiomass), scale))
  tryCatch({
    m1 <- lmer(LiveBiomass ~ PlotCwdMass * HostCwdMass * DeadHostBiomass * Elevation * Slope * Fire * PRam +
                    (1|SampleYear) + (1|Plot),
                    data=df1
                     ) 
    summary <- summary(m1)
    df_summary <- as.data.frame(summary$coefficients) |>
      mutate(Species = species,
              AIC = AIC(m1))
    df_summary$Term = rownames(df_summary)
    return(df_summary)
  }, error = function(e) {
    message("Error", e$message)
    return(NULL)
  })
}


#cleaning function for streamlined results
clean_me <- function(df) {
  aic_lookup <- df |> distinct(Species, AIC)
  df1 <- df |>
    rename(p = 5) |>
    mutate(Results = ifelse(
      p < 0.05,
      paste0("**", round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")**"),
      paste0(round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")")
    )) |>
    select(Species,Term,Results) |>
    pivot_wider(names_from="Term", values_from="Results") |>
    left_join(aic_lookup, by = "Species")
}


```

**This model uses all potential variables found to date with no interactions.**   

```{r}
add_models <- map_df(species_df, big_biomass_model) |>
  clean_me()
kable(add_models) 

```


**This model uses all potential variables found to date and allows for all possible interactions. Used to find interactions.**       

```{r}
int_models <- map_df(species_df, interaction_model) |>
  clean_me()
kable(int_models) 
```

```{r}
#| fig-width: 12 
#| fig-height: 14

int_me <- function(df,x,y,int) {
  df |>
    ggplot(aes(x={{x}},y={{y}},color=as.factor({{int}}),group=as.factor({{int}}))) +
      geom_point() +
      geom_smooth(method="lm") +
      climate_theme()
}

p1 <- int_me(biomass,DeadHostBiomass, HostCwdMass, PRam)
p2 <- int_me(biomass,DeadHostBiomass, Elevation, PRam)
p3 <- int_me(biomass,Fire, LiveBiomass,PRam)
p4 <- int_me(biomass,DeadHostBiomass, Slope, Fire)


(p1 + p2) / (p3 + p4)


```

**Other interactions to graph?**    

**This model uses an interaction between HostCwdMass, DeadHostBiomass, Fire and disease.**       

```{r}
disease_model <- function(species){
  df1 <- biomass |>
    filter(Species == {{species}}) |>
    mutate(across(c(PlotCwdMass, HostCwdMass, DeadHostBiomass, 
                  Elevation, Slope,LiveBiomass), scale))
  tryCatch({
    m1 <- lmer(LiveBiomass ~ PlotCwdMass + HostCwdMass * DeadHostBiomass * Fire * PRam + Elevation * Slope +
                    (1|SampleYear) + (1|Plot),
                    data=df1
                     ) 
    summary <- summary(m1)
    df_summary <- as.data.frame(summary$coefficients) |>
      mutate(Species = species,
              AIC = AIC(m1))
    df_summary$Term = rownames(df_summary)
    return(df_summary)
  }, error = function(e) {
    message("Error", e$message)
    return(NULL)
  })
}

disease_models <- map_df(species_df, disease_model) |>
  clean_me()
kable(disease_models) 

```

**How do the three models compare in AIC?**

```{r}
model_comp <- bind_rows(
  add_models |> mutate(Model = "Simple Model"),
  int_models |> mutate(Model = "Full Interaction Model"),
  disease_models |> mutate(Model = "Disease/Fire Interaction Model")
) |> 
  select(Model, Species, AIC) |>
  pivot_wider(names_from = Species, values_from = AIC) |>
  gt()

model_comp

```







### Logistic regression of mortality status  
-- Multiple logistic regression iterated across species   
    - Attempting to predict drivers of mortality given binary health status.  
- Improvements: Use mixed effects/survival analysis instead of logistic regression.   
**Questions I hope to answer**  
Which features, topographic or epidemiological, are driving mortality in Redwood forests?  
Is there a relationship between the explanatory variables and mortality?    

-- P-values for fixed variables below the 0.05 threshold will allow us to reject our null hypothesis and indicate a relationship between the explanatory variables and mortality.

```{r}
mortality <-  read_me("./clean/mortality_mixedmodel.csv")
species_mortality <- unique(mortality$Species)

mortality_model <- function(species){
  df1 <- mortality |>
    filter(Species == {{species}})
  logistic_model <- glm(Mortality ~ PlotCwdMass + HostCwdMass + Elevation + Slope + Fire + PRam,
                    data=df1, family = binomial
                     )
  summary <- summary(logistic_model)
  
  df_summary <- as.data.frame(summary$coefficients) |>
     mutate(Species = species)
  df_summary$Term = rownames(df_summary)
  return(df_summary)
}

mortality_results <- map_df(species_mortality, mortality_model)

mortality_results2 <- mortality_results |>
  rename(p = 4) |>
  mutate(Results = ifelse(
    p < 0.05,
    paste0("**", round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")**"),
    paste0(round(Estimate, 3), " (", formatC(p, format = "e", digits = 2), ")")
  )) |>
  select(Species,Term,Results) |>
  pivot_wider(names_from="Term", values_from="Results") |>
  select(!`(Intercept)`)

mortality_results3 <- mortality_results2 |>
  rename(Disease = PRam,
        `Plot Cwd Mass` = 2,
        `Host Cwd Mass` = 3)

kable(mortality_results3)

```

## Tracking stems through time   

### Survival Analysis

#### 1. Need to load and transform data into survival tables  
  - Create t and set t0 = 2006

```{r}
tree <- read_me("./clean/tree_table_FINAL.csv")

tree1 <- tree |>
  filter(Species %in% c("SESE", "LIDE","UMCA")) |> 
  rename(Disease = p_ram.presence,
         Fire = fire_presence) |>
  filter(!is.na(Species)) |>
  mutate(t = SampleYear - 2006,
         Species = case_when(
                             Species == "SESE" ~ "Redwood",
                             Species == "LIDE" ~ "Tanoak",
                             TRUE ~ "Bay laurel")) |>
  select(BS_plot, Species, t,RametFail,Disease,Fire, DBH)

```

#### [Kaplian - Meier](https://www.emilyzabor.com/survival-analysis-in-r.html#part-4-advanced-topics)  

```{r}
s1 <- survfit2(Surv(t,RametFail) ~ Species, data=tree1)

s1 |>
  ggsurvfit() +
  facet_wrap(~strata) +
  theme_classic() +
  labs(title="Kaplan-Meier Survival Function of Big Sur Species", x="Time")

#export_image(p1, "surv.png") 

summary(s1)
survdiff(Surv(t, RametFail) ~ Species, data = tree1)

``` 



### CWD Decomposition   
-- Influence of species,disease,fire on decomposition rate of CWD.    
-- Linearized exponential decay model iterated across the species.    
-- Aim to estimate values of k and compare between plot conditions.   
**Questions I hope to answer**  
How do fire, disease and species impact decomposition rates?   
Is there a linear relationship between change in mass and time?     

-- This piggy backs off earlier research conducted by Dr.Cobb. I am utilizing a known equation for [leaf litter decay](https://cdnsciencepub.com/doi/10.1139/x06-012). Results are simple linear regression, just a lot of build up.    
-- Improvements: Split the dataset into burned/unburned, run individual models for each


```{r}
cwd_decay <- read_me("./clean/cwd_decay.csv")
expo_decay <- function(species){
  df <- cwd_decay |>
    filter(Species == species) 
  model <- nls(Y ~ -k*t, data=df, start=list(k=0.7))
  df$PredictedY <- predict(model)
  summary <- summary(model)
  
  return(df)
}
# apply model function to all species
species_vec <- unique(cwd_decay$Species)
expo_results <- map_df(species_vec, expo_decay)
# need to prepare for visuals
expo_results2 <- expo_results |>
  mutate(Fire = as.factor(Fire)) |>
  drop_na() |>
  mutate(Burn = ifelse(Fire == 0, "Unburned", "Burned"))


expo_results2 |>
  ggplot(aes(x=t,y=Y, color=Burn)) +
  geom_point(shape=1, size=1.3) +
  geom_line(aes(x=t,y=PredictedY), color="red", linetype="dashed") +
  facet_wrap(~Species,scales = "free") + 
  labs(title="Linearized Exponential Decomposition Models for Studied Big Sur Trees", x="Time (t)", y="", subtitle="ln(Proportion of Mass Change in Year t)",
  color="Fire Status") +
  theme_classic()

```

### CWD Mixed Modeling  

```{r}


```



